\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-08-13 NAR_DRAFT.tex
fix TLOD
fix a bunch of p

At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is total read depth.
We want to use $\mathbf{x}$ to select between competing hypotheses;
\begin{equation}
  \label{eqn:hypothesis}
$$
  \begin{array}{l}
    \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
  \end{array}
$$
\end{equation}
where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
The posterior probability of a given hypothesis, $P(m,\nu)$, is the product of the likelihood of the data given the hypothesis and the prior probability of the hypothesis. 
Assuming that reads are independent, we have

@@

Assuming that the identity of the alternate allele and its allele frequency are independent of each other, and $\nu$ is uniformly distributed (i.e. $p(\nu) = 1$), we can rewrite equation \ref{eqn:1} as
\begin{equation}
  \label{eqn:2}
$$
  P(m,\nu) = p(m) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i).
$$
\end{equation}
Here we show how \batcave can be used to provide a tumor and site-specific estimate of the prior probability of mutation $p(m)$.

\subsection{Site-specific prior probability of mutation}
The probability which we have denoted $p(m)$ in equation \ref{eqn:2} for compactness is more precisely the joint probability that a mutation has occured $M$, and that it was to allele $m$, which we will denote here $p(m,M)$.
However, $p(m,M)$ is not constant for every site in the genome, and follows a distribution conditional on the genomic context surrounding the site, denoted here $p(m,M | C)$ \cite{Buisson2019}.
Assuming that $m$ and $M$ are independent conditional on the genomic context, $\mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(M \mid C)$, which we can use Bayes' rule to further decompose as

\begin{equation}
  \label{eqn:full_prior}
  $$
  \mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{p(M)}{p(C)}.
  $$
\end{equation}
We next show how to estimate the quantities in equation \ref{eqn:full_prior}.

@@

In order to estimate $\boldsymbol{\pi}$, we generate a subset of variants in the tumor in which we have high confidence that the variants are present, based on an initial calculation of their likelihood give the data.
These are variants such that for any reasonable value of the site-specific prior probability of mutation, the evidence for the variant in the read data erases any doubt about the presence of the variant. 
Let $\mathrm{D}$ be the set of high confidence variant calls, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
  $$
    \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\}.
  $$
Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C = c \mid M,\mathrm{D})$ as
\begin{equation}
  \label{eqn:post_pred}
  $$
  p(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
  $$
\end{equation}
The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C = c$, $p(m  \mid C = c,\mathrm{D})$ is then
\begin{equation}
  \label{eqn:to_allele}
  $$
   p(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
  $$
\end{equation}
The remaining context dependent component in equation \ref{eqn:detailed_prior} is $p(C = c)$, the prior probability of seeing a particular tri-nucleotide context.
We compute this as simply the proportion of all tri-nucleotide contexts with the context $C = c$ present in the target genome, which is slightly different depending on whether we are analyzing a whole genome or an exome.

@@

\begin{equation}
  \label{eqn:tlod}
    $$
    \mathrm{TLOD} = \frac{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = \hat{f}}(x_i)}{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = 0}(x_i)}.
    $$
\end{equation}

>>>>

At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is total read depth.
We want to use $\mathbf{x}$ to select between competing hypotheses;
\begin{equation}
  \label{eqn:hypothesis}
$$
  \begin{array}{l}
    \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
  \end{array}
$$
\end{equation}
where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
The posterior probability of a given hypothesis, $\mathrm{P}(m,\nu)$, is the product of the likelihood of the data given the hypothesis and the prior probability of the hypothesis. 
Assuming that reads are independent, we have

@@

Assuming that the identity of the alternate allele and its allele frequency are independent of each other, and $\nu$ is uniformly distributed (i.e. $\mathrm{p}(\nu) = 1$), we can rewrite equation \ref{eqn:1} as
\begin{equation}
  \label{eqn:2}
$$
  P(m,\nu) = p(m) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i).
$$
\end{equation}
Here we show how \batcave can be used to provide a tumor and site-specific estimate of the prior probability of mutation $p(m)$.

\subsection{Site-specific prior probability of mutation}
The probability which we have denoted $\mathrm{p}(m)$ in equation \ref{eqn:2} for compactness is more precisely the joint probability that a mutation has occured $M$, and that it was to allele $m$, which we will denote here $\mathrm{p}(m,M)$.
However, $\mathrm{p}(m,M)$ is not constant for every site in the genome, and follows a distribution conditional on the genomic context surrounding the site, denoted here $\mathrm{p}(m,M | C)$ \cite{Buisson2019}.
Assuming that $m$ and $M$ are independent conditional on the genomic context, $\mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(M \mid C)$, which we can use Bayes' rule to further decompose as

\begin{equation}
  \label{eqn:full_prior}
  $$
  \mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C)}.
  $$
\end{equation}
We next show how to estimate the quantities in equation \ref{eqn:full_prior}.

@@

In order to estimate $\boldsymbol{\pi}$, we generate a subset of variants in the tumor in which we have high confidence that the variants are present, based on an initial calculation of their likelihood give the data.
These are variants such that for any reasonable value of the site-specific prior probability of mutation, the evidence for the variant in the read data erases any doubt about the presence of the variant. 
Let $\mathrm{D}$ be the set of high confidence variant calls, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
  $$
    \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\}.
  $$
Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C = c \mid M,\mathrm{D})$ as
\begin{equation}
  \label{eqn:post_pred}
  $$
  \mathrm{p}(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
  $$
\end{equation}
The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C = c$, $p(m  \mid C = c,\mathrm{D})$ is then
\begin{equation}
  \label{eqn:to_allele}
  $$
   \mathrm{p}(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
  $$
\end{equation}
The remaining context dependent component in equation \ref{eqn:detailed_prior} is $p(C = c)$, the prior probability of seeing a particular tri-nucleotide context.
We compute this as simply the proportion of all tri-nucleotide contexts with the context $C = c$ present in the target genome, which is slightly different depending on whether we are analyzing a whole genome or an exome.

@@

\begin{equation}
  \label{eqn:tlod}
    $$
    \mathrm{TLOD} = \mathrm{log}_{10}\left(\frac{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = \hat{f}}(x_i)}{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = 0}(x_i)}\right).
    $$
\end{equation}
